### 2015-08-26 LaTeX/RhodiumEmp.tex
fixed typos and stuff

We collect information on the location of conflicts from the PRIO Conflict Site Dataset \citep{hallberg:2012}. This dataset contains information on the geographical centers of armed conflict events and the area covered by each conflict at a yearly level from 1989 to 2008. In recent years, other geo-referenced datasets have been developed that go beyond just providing information on conflict centroids and area to providing an actual spatial grid structure through which to conduct analysis \citep{tollefsen:etal:2012}. However, these data are not available globally and would therefore unnecessarily limit the external validity of the current analysis.\footnote{Nonetheless, we conduct a robustness test of our primary models using data from the Armed Conflict Location and Event Data (ACLED) Project. The results described here hold with the new data. A description of the ACLED modeling strategy and results can be found in the Appendix.} Other machine coded event datasets, while providing detailed coverage of events on a global scale, often approximate event locations based on descriptors in the source material or by the location of the reporting news agency itself. Because the location of conflict is critical for this analysis, the urban bias of disaggregated event data is unacceptable. For the purpose of this analysis, we deliberately avoid using these datasets since our primary interest is in examining variation in changes to GDP growth at a national level.

To construct a time-series cross-sectional database of urban centers, we turn to yearly editions of \emph{The World Alamanac}. More refined subnational data listing urban centers by their contribution to economic output would be the most direct way to test our hypothesis, however, cross-national data such as this over the time period of our analysis, especially for developing countries, is simply not available. \emph{The World Alamanac} lists the major cities in a country by population, thus providing us with a second best approximation of the urban centers that are most relevant for a particular national economy. Typically, the \emph{Almanacs} list at least three major cities, including the capital, for each country and year from 1989 to 2008. Because the \emph{Almanacs} are not perfectly consistent from year to year, we opted to code cities with a five year rolling window. That is, cities listed in the 1995 \emph{Almanac} are also included in our dataset for years 1993-1997. In doing so, we hope to minimize the effect of coding inconsistencies on the part of \emph{The World Almanac}. Also, as a more conservative test of our hypothesis, we replicate our models using the distance of conflict from each capital city rather than our preferred measure of minimum distance to a major urban center. While our hypothesis is most accurately represented by distance from any major urban centers, distance from capital cities provides an appealing and consistent alternative measure. Every country has a capital city in our sample, so using this alternative measure does not rely on differing population or density thresholds across countries that must be used in coding major cities.

In figure \ref{fig:CityConfMap}, we show the geographic distribution of conflicts and cities. The centroid locations of conflicts are shown by red dots. Not surprisingly, in many cases conflict locations are clustered within specific parts of a country. In most cases, this clustering is indicative of the same conflict moving within the geographic boundaries of a country over time. The blue diamonds in \ref{fig:CityConfMap} denote the locations of major cities from 1989 to 2008. Countries shaded in grey are those for which no armed conflict took place in this period according to the PRIO dataset.

@@

Using our geo-referenced conflict and urban center data, we proceed to compute the proximity of every civil armed conflict in a country-year from every urban center listed in that year.\footnote{Distances between centroid locations were computed using an iterative method of distance calculation proposed by \citet{vincenty:1975}.} Since our unit of analysis is the country-year, we aggregate the distance between conflicts and major cities by calculating the minimum logged distance any conflict is from a major city. For example, if a country faces four conflicts in a year, the datapoint for that country-year would be the minimum distance any conflict centroid is from any city centroid. In addition to identifying the minimum distance a conflict is from any city, we also create a variable that measures minimum distance of any conflict from the capital of the country. Given the importance of cities as drivers of macroeconomic growth the proximity of conflict to even one city should have at least a noticeable short-term impact on growth. Thus this choice of aggregation conforms clostest to our theoretical claims about how macroeconomic growth can be severely disrupted in cases where conflicts are proximate to major cities.

Figure \ref{fig:distGdp} provides a cursory illustration of the relationship between GDP growth and our spatial proximity variables. Each bar on the rightmost panel represents the average GDP growth rate across country-years for which the minimum distance between the centroid of the conflict and capital city fell within a certain quartile range across our full sample. Clearly, as the minimum distance from a conflict to capital city declines, there is a sharp decline in average growth rates for the following year. On the rightmost panel of this figure, the same relationship holds when we test the effect of the minimum distance of conflict from any major urban center.

@@

Next, we move beyond this cursory illustration and explicate our full modeling strategy. Given how we constructed our spatial proximity of conflict variables, it would make little sense to include both within one model. The obvious problem we would run into is that there would be cases for which the minimum distance of a conflict to any city would be the same as the minimum distance to the capital. Thus we estimate two separate models, one where we include the minimum distance from a conflict to the capital and the other where we include the minimum distance to any urban center. In addition to the inclusion of our spatial proximity covariates we include a number of control measures to account for extant explanations of how civil war dynamics affect GDP growth and to mitigate omitted variable bias.

From the PRIO Armed Conflict dataset we bring in additional information about conflict intensity, duration, the area covered by the conflict, and the number of conflicts \citep{themner:wallensteen:2014}. Conflict intensity is a binary that distinguishes between minor armed conflicts (between 25 and 999 annual battle deaths) and wars (at least 1000 annual battle deaths), and we set this equal to one if any of the conflicts in a country year were classified as a war and zero if not. In aggregating conflict duration to the country year level, we bring in the duration (measured in years) of the longest lasting conflict by a certain country-year. Conflict area measures the conflict zone in square kilometers. To include this in our country-year model, we sum up the area covered by each conflict being fought and divide by the total land area of the country. Finally, the number of conflicts is simply a count of the number of ongoing internal armed conflicts in a country-year according to the PRIO dataset.

@@

Aggregating each of the PRIO conflict descriptive measures in this way is to help account for those countries in which conflicts are especially intense, cover large areas, have been long-lasting, or are one of many ongoing conflicts. Given the findings in the extant literature, our expectation is that each of these measures should have a negative effect on economic growth. Though we also expect that the substantive implications of any one of these measures will matter less than the adverse effect to economic growth by the spatial proximity of conflict to urban centers.

We also include a number of variables that help to capture structural and institutional aspects of the country itself. First, we include a binary for whether the country is classified as upper income by the World Bank and for the level of inflation in the country. Second, we also control for the level of democracy in the country using the polity index \citep{marshall:etal:2013}. We also control for the proportion of a country's GDP that is made up of natural resource rents. This is a particularly important variable to control for as countries that draw a large share of their GDP from natural resource rents likely rely less on urban centers to serve as drivers of growth. Finally, given that the timeframe of our sample overlaps with the occurrence of two major global economic crises, the 1997 Asian financial crisis and the 2008 financial crisis, we add a control for the average GDP growth across all countries in the world.

To estimate variation in growth rates, we use a random effects model clustered on countries. A random effects specification is preferred to using fixed effects here as we are not interested in estimating the change in GDP within units over time, rather our purpose is to explain variation between units. However, to ensure that it is also statistically justifiable to make this choice we run a Hausman specification test \citep{greene:2008}. For both the minimum conflict distance to any and to capital city specifications we fail to reject the null hypothesis at either the 90 and 95\% confidence intervals, providing at least some initial evidence that we are justified in our choice. However, as \citet{clark:linzer:2015} note the Hausman test should not be the sole determination for choosing between fixed or random effects. We utilize their more nuanced typology for further arbitrating between fixed and random effects, and again we find that random effects are preferred in this case.\footnote{Specifically to take into consideration the size of the dataset (both number of units, in this case countries, and number of observations per unit, conflict instances) and the level of correlation between the regressor and unit effects. In our case, we have over 70 countries but for over half of those we only have five conflict instances or less. Given such a data structure Clark \& Linzer (2015) recommend to examine the level of correlation between the regressor and unit effects to determine the appropriate modeling framework. For both our distance models, the level of correlation between the regressor and unit effects is less than 0.20, which accords with a random effects recommendation under the framework described by Clark \& Linzer (2015).}

The total number of observations we have for this analysis are 505 country-year cases of civil war from 68 different countries during the period of 1989 to 2008. The estimation model is shown below, where $Ln(Min. \; Conflict \; Dist.)_{i,t-1}$ is a placeholder for $Ln(Min. \; Capital \; Dist.)_{i,t-1}$ and $Ln(Min. \; City \; Dist.)_{i,t-1}$:

@@

Before we move to discussing the results of the model specification that we have described above we estimate a simple, baseline model to assess the effect of any civil conflict on growth. For this model we use a specification as similar as possible to the one we explicated in the previous section, with the key difference being that this baseline model is estimated using a full country-year panel with fixed effects.\footnote{In choosing between fixed and random effects, we again began by running a Hausman test and followed the recommendations set out in \citet{clark:linzer:2015}. In this case, found that the fixed effects specification was preferred.} The dependent variable is again annual percent change in GDP from the World Development Indicators, and the key independent variable here is whether or not the country was engaged in a civil war in the previous year according to the PRIO Armed Conflict dataset. Since we are using a full country-year panel, we have to omit the PRIO conflict specific variables (e.g., conflict intensity, conflict duration) as they are undefined for country-year observations in which no conflict was taking place. The rest of the independent variables are similar to the model specified in the previous section, specifically, we include: whether or not the country is classified as upper income, lagged version of inflation, lagged version of polity, lagged version of resource rents as a proportion of GDP, and average GDP growth across all countries in the world.

@@

We will eschew discussing the controls since the primary purpose of this analysis is to determine a baseline effect for civil war. Additionally, for the sake of space we do not show the full results from this regression, but they are available in section \ref{tab:baseline} of the Appendix. The key result pertains to the civil war variable, the $\beta$ estimate for this variable is approximately -2.568 with a standard error of 0.502.

To assess the substantive significance of this variable we turn to a simulation based analysis. To to do this we set up two scenarios, one in which a civil war occurred and in the other no civil war, all the other parameters were set to their median value. Next, we conduct 1,000 random draws from a multivariate normal to obtain a distribution of point estimates of each regression coefficient. Last, we simply matrix multiply the draws from the multivariate normal with the transposed scenario matrix to obtain expected values of GDP growth by the two scenarios. The resulting distributions are shown in figure \ref{fig:civWarEffect}.

@@

The analysis here highlights that there is little distinguishable difference in the expected values of GDP growth by whether or not there a country was experiencing a civil war in the previous year. We argue that this is because the effect of civil conflict on economic growth becomes particularly relevant when that conflict is proximate to economic points of interest such as capital cities and major urban centers.

@@

Turning to our hypothesis, we find strong support for the argument that spatial proximity of conflict to either a capital city or any major urban center is associated with lower levels of GDP growth. The positive coefficient on the logged, minimum conflict distance variables indicates that conflicts closer to major cities have an adverse effect on economic growth.

@@

To assess the substantive effect of these results, we conduct a number of simulations. We set up scenarios where we hold all variables to their median except for logged, minimum conflict distance, which we range from its minimum to maximum value. Next, we conduct 1,000 random draws from a multivariate normal to obtain distributions for the point estimates of each of the regression coefficients. After obtaining these distributions, we calculate the predicted value of GDP growth based on the conditions set by the scenarios. We plot the results of this analysis in figures \ref{fig:capSim} and \ref{fig:citySim}.

@@

The simulation results for both $Ln(Min. \; Capital \; Dist.)_{i,t-1}$ and $Ln(Min. \; City \; Dist.)_{i,t-1}$ are quite similar. In both cases, conflicts located farther away from major cities have almost no adverse impact on economic growth, and, in fact, are likely to still see positive growth. As the proximity of conflicts to urban centers increases, however, we can clearly see substantive declines in growth rates start to emerge. Thus, through utilizing these spatial proximity measures we can start to explain why there is such meaningful variation in economic performance amongst countries facing internal armed conflict.

>>>>

We collect information on the location of conflicts from the PRIO Conflict Site Dataset \citep{hallberg:2012}. This dataset contains information on the geographical centers of armed conflict events and the area covered by each conflict at a yearly level from 1989 to 2008. In recent years, other geo-referenced datasets have been developed that go beyond just providing information on conflict centroids and area to providing an actual spatial grid structure representing high-resolution spatial distributions of conflict \citep{tollefsen:etal:2012}. However, these data are not available globally and would therefore unnecessarily limit the external validity of the current analysis.\footnote{Nonetheless, we conduct a robustness test of our primary models using data from the Armed Conflict Location and Event Data (ACLED) Project. The results described here hold with the new data. A description of the ACLED modeling strategy and results can be found in the Appendix.} Other machine coded event datasets, while providing detailed coverage of events on a global scale, often approximate event locations based on descriptors in the source material or by the location of the reporting news agency itself. Because the location of conflict is critical for this analysis, the urban bias of disaggregated event data is unacceptable.

To construct a time-series cross-sectional database of urban centers, we turn to yearly editions of \emph{The World Alamanac}. More refined subnational data listing urban centers by their contribution to economic output would be the most direct way to test our hypothesis; however, cross-national data such as this over the time period of our analysis, especially for developing countries, is simply not available. \emph{The World Alamanac} lists the major cities in a country by population, thus providing us with a second best approximation of the urban centers that are most relevant for a particular national economy. Typically, the \emph{Almanacs} list at least three major cities, including the capital, for each country and year from 1989 to 2008. Because the \emph{Almanacs} are not perfectly consistent from year to year, we opted to code cities with a five year rolling window. That is, cities listed in the 1995 \emph{Almanac} are also included in our dataset for years 1993-1997. In doing so, we hope to minimize the effect of coding inconsistencies on the part of \emph{The World Almanac}. Also, as a more conservative test of our hypothesis, we replicate our models using the distance of conflict from each capital city rather than our preferred measure of minimum distance to a major urban center. While our hypothesis is most accurately represented by distance from any major urban centers, distance from capital cities provides an appealing and consistent alternative measure. Every country has a capital city in our sample so using this alternative measure does not rely on differing population or density thresholds across countries that must be used in coding major cities.

In figure \ref{fig:CityConfMap}, we show the geographic distribution of conflicts and cities. The centroid locations of conflicts are shown by red dots. Not surprisingly, in many cases conflict locations are clustered within specific parts of a country. In most cases, this clustering is indicative of the same conflict moving within the geographic boundaries of a country over time. The blue diamonds in figure \ref{fig:CityConfMap} denote the locations of major cities from 1989 to 2008. Countries shaded in grey are those for which no armed conflict took place in this period according to the PRIO dataset.

@@

Using our geo-referenced conflict and urban center data, we proceed to compute the proximity of every domestic armed conflict in a country-year to every urban center listed in that year.\footnote{Distances between centroid locations were computed using an iterative method of distance calculation proposed by \citet{vincenty:1975}.} Since our unit of analysis is the country-year, we aggregate the distance between conflicts and major cities by calculating the minimum logged distance any conflict is from a major city. For example, if a country faces four conflicts in a year, the datapoint for that country-year would be the minimum distance any conflict centroid is from any city centroid. In addition to identifying the minimum distance a conflict is from any city, we also create a variable that measures minimum distance of any conflict from the capital of the country. Given the importance of cities as drivers of macroeconomic growth, the proximity of conflict to even one city should have at least a noticeable short-term impact on growth. Thus this choice of aggregation conforms closest to our theoretical claims about how macroeconomic growth can be severely disrupted in cases where conflicts are proximate to major cities.\footnote{We ran robustness tests using an alternative aggregation method for conflict distance. Rather than selecting the minimum distance of any conflict to any city in a given year, we took the average minimum distance of each conflict to the nearest city. The results were statistically significant and substantively similar across parameterizations.}

Figure \ref{fig:distGdp} provides a cursory illustration of the relationship between GDP growth and our spatial proximity variables. Each bar on the leftmost panel represents the average GDP growth rate across country-years for which the minimum distance between the centroid of the conflict and capital city fell within a certain quartile range across our full sample. Clearly, as the minimum distance from a conflict to capital city declines, there is a sharp decline in average growth rates for the following year. On the rightmost panel of this figure, the same relationship holds when we test the effect of the minimum distance of conflict from any major urban center.

@@

Next, we move beyond this cursory illustration and explicate our full modeling strategy. Given how we constructed our spatial proximity of conflict variables, it would make little sense to include both capital distance and city distance within one model. The obvious problem we would run into is that there would be cases for which the minimum distance of a conflict to any city would be the same as the minimum distance to the capital. Thus we estimate two separate models, one where we include the minimum distance from a conflict to the capital and the other where we include the minimum distance to any urban center. In addition to the inclusion of our spatial proximity covariates we include a number of control measures to account for extant explanations of how civil war dynamics affect GDP growth and to mitigate omitted variable bias.

From the PRIO Armed Conflict dataset we bring in additional information about conflict intensity, duration, the area covered by the conflict, and the number of conflicts \citep{themner:wallensteen:2014}. Conflict intensity is a binary that distinguishes between minor armed conflicts (between 25 and 999 annual battle deaths) and wars (at least 1000 annual battle deaths), and we set this equal to one if any of the conflicts in a country year were classified as a war and zero if not.\footnote{We also conduct robustness tests in which high intensity and low intensity subsets of the data are modeled separately. These can be found in Table \ref{tab:modHiLoIntensity} of the Appendix.} In aggregating conflict duration to the country year level, we bring in the duration (measured in years) of the longest lasting conflict by a certain country-year. Conflict area measures the conflict zone in square kilometers. To include this in our country-year model, we sum up the area covered by each conflict being fought and divide by the total land area of the country. Finally, the number of conflicts is simply a count of the number of ongoing internal armed conflicts in a country-year according to the PRIO dataset.

@@

Aggregating each of the PRIO conflict descriptive measures in this way is to help account for those countries in which conflicts are especially intense, cover large areas, have been long-lasting, or are one of many ongoing conflicts. Given the findings in the extant literature, our expectation is that each of these measures should have a negative effect on economic growth. Though we also expect that the substantive implications of any one of these measures will matter less than the adverse effect to economic growth accounted for by the spatial proximity of conflict to urban centers.

We also include a number of variables that help to capture structural and institutional aspects of the country itself. First, we include binary indicators for whether the country is classified as upper income by the World Bank and for the level of inflation in the country. Second, we control for the level of democracy in the country using the polity index \citep{marshall:etal:2013}. We also control for the proportion of a country's GDP that is made up of natural resource rents. This is a particularly important variable to control for as countries that draw a large share of their GDP from natural resource rents likely rely less on urban centers to serve as drivers of growth. Finally, given that the time frame of our sample overlaps with the occurrence of two major global economic crises, the 1997 Asian financial crisis and the 2008 financial crisis, we add a control for the average GDP growth across all countries in the world.

To estimate variation in growth rates, we use a random effects model clustered on countries. A random effects specification is preferred to using fixed effects here as we are not interested in estimating the change in GDP within units over time; rather our purpose is to explain variation between units. However, to ensure that it is also statistically justifiable to make this choice we run a Hausman specification test \citep{greene:2008}. For both the minimum conflict distance to any and to capital city specifications we fail to reject the null hypothesis at both the 90 and 95\% confidence intervals, providing at least some initial evidence that we are justified in our choice. However, as \citet{clark:linzer:2015} note, the Hausman test should not be the sole determinant for choosing between fixed or random effects. We utilize their more nuanced typology for further arbitrating between fixed and random effects, and again we find that random effects are preferred in this case.\footnote{We take into consideration the size of the dataset (both number of units, in this case countries, and number of observations per unit, conflict instances) and the level of correlation between the regressor and unit effects. In our case, we have over 70 countries but for over half of those we only have five conflict instances or less. Given such a data structure, Clark \& Linzer (2015) recommend to examine the level of correlation between the regressor and unit effects to determine the appropriate modeling framework. For both our distance models, the level of correlation between the regressor and unit effects is less than 0.20, which accords with a random effects recommendation under the framework described by Clark \& Linzer (2015).}

For this analysis we have 505 country-year cases of civil war from 68 different countries during the period between 1989 to 2008. The model formula (omitting error terms) is shown below, where $Ln(Min. \; Conflict \; Dist.)_{i,t-1}$ is a placeholder for $Ln(Min. \; Capital \; Dist.)_{i,t-1}$ and $Ln(Min. \; City \; Dist.)_{i,t-1}$:

@@

Before we move to discussing the results of the model specification that we have described above, we estimate a simple baseline model to assess the effect of any civil conflict on growth. For this model we use a specification as similar as possible to the one we explicated in the previous section, with the key difference being that this baseline model is estimated using a full country-year panel with fixed effects.\footnote{In choosing between fixed and random effects, we again began by running a Hausman test and followed the recommendations set out in \citet{clark:linzer:2015}. In this case we found that the fixed effects specification was preferred.} The dependent variable is again annual percent change in GDP from the World Development Indicators, and the key independent variable here is whether or not the country was engaged in a civil war in the previous year according to the PRIO Armed Conflict dataset. Since we are using a full country-year panel, we have to omit the PRIO conflict specific variables (e.g., conflict intensity, conflict duration) as they are undefined for country-year observations in which no conflict was taking place. The rest of the independent variables are similar to the model specified in the previous section. Specifically, we include: whether or not the country is classified as upper income, the lagged version of inflation, the lagged version of polity, the lagged version of resource rents as a proportion of GDP, and the average GDP growth across all countries in the world.

@@

We will eschew discussing the controls since the primary purpose of this analysis is to determine a baseline effect for civil war. Additionally, for the sake of space we do not show the full results from this regression, but they are available in section \ref{tab:baseline} of the Appendix. The key result pertains to the civil war variable for which we estimate a regression coefficient of approximately -2.568 with a standard error of 0.502.

To assess the substantive significance of this variable we turn to a simulation-based analysis. We simulate two scenarios, one in which a civil war occurred and another in which there is no civil war.  All of the other parameters are set to their median value. Next, we conduct 1,000 random draws from a multivariate normal distribution to obtain a distribution of point estimates for each regression coefficient. Last, we simply matrix multiply the draws from the multivariate normal with the transposed scenario matrices to obtain expected values of GDP growth given the two scenarios. The resulting distributions are shown in figure \ref{fig:civWarEffect}.

@@

The analysis here highlights that there is little distinguishable difference in the expected values of GDP growth as a function of whether or not there a country was experiencing a civil war in the previous year. We argue that this is because the effect of civil conflict on economic growth becomes particularly relevant when that conflict is proximate to economic points of interest such as capital cities and major urban centers.

@@

Turning to our hypothesis, we find strong support for the argument that spatial proximity of conflict to either a capital city or any major urban center is associated with lower levels of GDP growth. The positive coefficient on the logged minimum conflict distance variables indicates that conflicts closer to major cities have an adverse effect on economic growth.

@@

To assess the substantive effect of these results, we conduct a number of simulations. We set up scenarios where we hold all variables to their medians except for logged minimum conflict distance, which we range from its minimum to maximum value. Next, we conduct 1,000 random draws from a multivariate normal to obtain distributions for the point estimates of each of the regression coefficients. After obtaining these distributions, we calculate the predicted value of GDP growth based on the conditions set by the scenarios. We plot the results of this analysis in figures \ref{fig:capSim} and \ref{fig:citySim}.

@@

The simulation results for both $Ln(Min. \; Capital \; Dist.)_{i,t-1}$ and $Ln(Min. \; City \; Dist.)_{i,t-1}$ are quite similar. In both cases, conflicts located farther away from major cities have almost no adverse impact on economic growth, and, in fact, are likely to still see positive growth. As the proximity of conflicts to urban centers increases, however, we can clearly see substantive declines in growth rates. Thus, through utilizing these spatial proximity measures, we can begin to explain why there is such meaningful variation in economic performance amongst countries facing internal armed conflict.

### 2015-08-26 LaTeX/RhodiumLit.tex
recompiled with new intro
fixed typos and stuff

Instead of disaggregating economic outcomes, \citet{imai:weinstein:2000} instead disaggregate conflict itself. They distinguish between those conflicts that cover larger or smaller geographic areas and hypothesize that larger conflicts (in terms of geographical spread) will result in worse economic performance. Using a variety of regression techniques, they find that there is a negative correlation between the geographical spread of conflict and the decade average of economic growth for each country. Widespread conflicts, they argue, are more likely to result in damage to infrastructure, divestment from normal state spending, and capital flight. Their results suggest that widespread civil wars are five times more costly than those fought over a narrower geographic area and can reduce GDP grotwh by 1.25\% annually.

@@

In accordance with this theory, \citet{fearon:laitin:2003} argue that poor economic growth is the primary condition conducive for civil war. More specifically, they believe that strong economic growth proxies for robust governance and that states with low GDP growth likely have infrastructures that are unable to implement counter-insurgent policies. In an effort to parse out the causal effect of economic growth shocks on civil war, \citet{miguel:etal:2004} instrument income growth with rainfall. They find that rainfall is strongly correlated with income in sub-Saharan Africa, a region also prone to civil conflict in recent decades.  Using a two-stage estimation approach, they conclude income is correlated with the likelihood of civil war.

>>>>

Instead of disaggregating economic outcomes, \citet{imai:weinstein:2000} instead disaggregate conflict itself. They distinguish between those conflicts that cover larger or smaller geographic areas and hypothesize that larger conflicts (in terms of geographical spread) will result in worse economic performance. Using a variety of regression techniques, they find that there is a negative correlation between the geographical spread of conflict and the decade average of economic growth for each country. Widespread conflicts, they argue, are more likely to result in damage to infrastructure, divestment from normal state spending, and capital flight. Their results suggest that widespread civil wars are five times more costly than those fought over a narrower geographic area and can reduce GDP growth by 1.25\% annually.

@@

In accordance with this theory, \citet{fearon:laitin:2003} argue that poor economic growth is the primary condition conducive for civil war. More specifically, they believe that strong economic growth proxies for robust governance and that states with low GDP growth likely have infrastructures that are unable to implement counter-insurgent policies. In an effort to parse out the causal effect of economic shocks on civil war, \citet{miguel:etal:2004} instrument income growth with rainfall. They find that rainfall is strongly correlated with income in sub-Saharan Africa, a region also prone to civil conflict in recent decades.  Using a two-stage estimation approach, they conclude that income is correlated with the likelihood of civil war.

### 2015-08-26 LaTeX/RhodiumTheory.tex
fixed typos and stuff

In the past three decades, as the pace of urbanization has accelerated, economists have devoted even more energy to investigating the role of city-level economic drivers \citep{lucas:1988, ciccone:hall:1996, begg:1999, henderson:wang:2007}. Most notably, \citeauthor{krugman:1991a} in his seminal 1991 article, ``Increasing Returns and Economic Geography'', convincingly showed the importance of studying economic outcomes within a spatial context. A key finding of Krugman's work on economic geography has been the importance of agglomeration effects, the idea that ``activities tend to cluster where markets are large and markets become larger where activities cluster'' \citep{krugman:1997}. \citet{henderson:2000} provides empirical evidence for why these agglomeration effects would manifest by showing that spatial clustering promotes economic efficiency in a variety of ways from simply enabling savings in transport costs to more efficiently functioning labor markets.

In addition to agglomeration effects, there is an even simpler story, which can be traced back to \citet{marshall:1920}, that states geographic clustering of firms promotes valuable learning and exchange between actors. \citet{lucas:1988} provides a formal treatment of this argument and shows that the accumulation of human capital generates positive spillovers, where if even one worker acquires a new skill then spatially proximate workers would all become more productive. Empirical estimates of agglomeration effects indicate that a doubling of employment density correspond to a 5\% increase in labor productivity \citep{ciccone:hall:1996,ciccone:2002}. \citet{glaeser:etal:1992} examine this problem empirically and show that cities are important for economic growth precisely because they lead to the knowledge spillovers that are important for innovation. Similar empirical results have been found by many others, which has led to a growing consensus on how cities can serve as important knowledge hubs for a national economy \citep{jaffe:etal:1993, glaeser:1994, firestone:2010}.

@@

\citet{glaeser:shapiro:2002} discuss the historic role of cities in warfare and their current role with respect to terrorism. Cities, they argue, once provided safe haven in the midst of conflict; defense of a concentrated population is easier than defense of a dispersed population. However, as the tactics of warfare have shifted over the centuries, the relationship between violent conflict and cities has become more complex. Their dense populations also tempt belligerents and maximize the impact to cost ratio of a given violent action.

Additionally, warfare destroys transportation infrastructure, which can interfere both with commerce and with rebuilding during a conflict. In particular, conflict affects citizens ability and willingness to participate in commerce. When major population centers are threatened by violence, residents will be less likely to engage in economically productive activities. Violence near major population centers not only threatens residents directly, but impedes business by threatening trade between the population center and other cities or rural areas.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=.5\textwidth]{colombiaMap-crop}
	\caption{This map illustrates the geographic distribution of all conflicts in Colombia, according to the PRIO Conflict Site Dataset, and major cities from 1989 to 2008.}
	\label{fig:columbiaMap}
\end{figure}

@@

While FARC seems to exploit its ability to target areas of economic importance including cities, other insurgencies tend to be more peripheral. India, for instance, has faced challenges by armed groups in its north-east for half a century. However, this area of India is remote, primarily agrarian, and relatively less populous than other parts of India. Indeed, for much of this period, India has experienced relatively robust economic growth. In figure \ref{fig:indiaMap}, we show the geographic distribution of conflict in India from 1989 to 2007 again using the PRIO conflict site database. The story from this map is clearly quite stark from that of Colombia. Whereas in Colombia conflict had come right to the gates of major cities, in India conflict clearly has been confined to the periphery.

\begin{figure}[ht]
	\centering
	\includegraphics[width=.5\textwidth]{indiaMap-crop}
	\caption{This map illustrates the geographic distribution of all conflicts in India, according to the PRIO Conflict Site Dataset, and major cities from 1989 to 2007. }
	\label{fig:indiaMap}
\end{figure}

Nigeria and Cameroon have battled Boko Haram in a bloody conflict in their rural northern regions for half a decade. For the period between 2009 and 2013, the National Consortium for the Study of Terrorism and Responses to Terrorism (START) described Boko Haram as ``among the deadliest [terrorist groups] in the world'' \citep{pate:etal:2014}. The groups tactics range from car and suicide bombings to direct assaults to mass kidnappings. Nonetheless, Nigeria and Cameroon have both maintained strong economic growth during this period. The African Development Bank Group (AFDB) described Nigeria's economic growth from 2004 to 2014 as ``robust'' and projected ``moderate'' growth of 5\% for 2015 \citep{afdb:nigeria:2015}. On Cameroon, the AFDB said ``despite the security and humanitarian crisis in the region, Cameroonian growth remains strong at above 5\%'' \citep{afdb:cameroon:2015}.

It is important to note that Mexico does not appear in the civil war dataset chosen for this study. The UCDP definition of civil war requires that the conflict be fought over an ``incompatibility that concerns government and/or territory,'' a criteria not strictly met by the organized crime violence experienced by the country \citep[p.1]{themner:2014}. Nonetheless, we want to reiterate that this instance of armed conflict, of the magnitude typically associated with civil war, conforms to the expectations of our theory. Mexico, now nearly a decade into a violent and complicated conflict between several organized criminal enterprises and the federal government, has maintained healthy economic performance. For much of this time, the cartel violence generally occurred in rural areas along drug trafficking routes and not within major cities. \citet{beittel:2011} writes in 2011 that ``drug trafficking-related killings remain concentrated in a relatively few cities.'' Meanwhile, several majors cities including the capital, Mexico City, have experienced relatively low levels of cartel violence.

>>>>

In the past three decades, as the pace of urbanization has accelerated, economists have devoted even more energy to investigating the role of city-level economic drivers \citep{lucas:1988, ciccone:hall:1996, begg:1999, henderson:wang:2007}. Most notably \citeauthor{krugman:1991a}, in his seminal 1991 article, ``Increasing Returns and Economic Geography'', convincingly showed the importance of studying economic outcomes within a spatial context. A key finding of Krugman's work on economic geography has been the importance of agglomeration effects, the idea that ``activities tend to cluster where markets are large and markets become larger where activities cluster'' \citep{krugman:1997}. \citet{henderson:2000} provides empirical evidence for why these agglomeration effects would manifest by showing that spatial clustering promotes economic efficiency in a variety of ways from simply enabling savings in transport costs to producing more efficiently functioning labor markets.

In addition to agglomeration effects, there is an even simpler story, which can be traced back to \citet{marshall:1920}, that states geographic clustering of firms promotes valuable learning and exchange between actors. \citet{lucas:1988} provides a formal treatment of this argument and shows that the accumulation of human capital generates positive spillovers, where if even one worker acquires a new skill then spatially proximate workers would all become more productive. Empirical estimates of agglomeration effects indicate that a doubling of employment density corresponds to a 5\% increase in labor productivity \citep{ciccone:hall:1996,ciccone:2002}. \citet{glaeser:etal:1992} examine this problem empirically and show that cities are important for economic growth precisely because they lead to the knowledge spillovers that are important for innovation. Similar empirical results have been found by many others and have led to a growing consensus that cities can serve as important knowledge hubs for national economies \citep{jaffe:etal:1993, glaeser:1994, firestone:2010}.

@@

\citet{glaeser:shapiro:2002} discuss the historic role of cities in warfare and their current role with respect to terrorism. Cities, they argue, once provided safe haven in the midst of conflict; defense of a concentrated population is easier than defense of a dispersed population. However, as the tactics of warfare have shifted over the centuries, the relationship between violent conflict and cities has become more complex. Dense populations also tempt belligerents and maximize the impact to cost ratio of a given violent action.

Additionally, warfare destroys transportation infrastructure, which can interfere both with commerce and with rebuilding during a conflict. In particular, conflict affects a citizen's ability and willingness to participate in commerce. When major population centers are threatened by violence, residents will be less likely to engage in economically productive activities. Violence near major population centers not only threatens residents directly, but impedes business by threatening trade between the population center and other cities or rural areas.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=.5\textwidth]{colombiaMap-crop}
	\caption{This map illustrates the geographic distribution of all conflict centroids in Colombia, according to the PRIO Conflict Site Dataset, and major cities from 1989 to 2008.}
	\label{fig:columbiaMap}
\end{figure}

@@

While FARC seems to exploit its ability to target areas of economic importance including cities, other insurgencies tend to be more peripheral. India, for instance, has faced challenges by armed groups in its north-east for half a century. However, this area of India is remote, primarily agrarian, and relatively less populous than other parts of India. Indeed, for much of this period, India has experienced relatively robust economic growth. In figure \ref{fig:indiaMap}, we show the geographic distribution of conflict in India from 1989 to 2007 again using the PRIO conflict site database. The story from this map is clearly quite stark from that of Colombia. Whereas in Colombia conflict had come right to the gates of major cities, in India conflict has been primarily confined to the periphery.

\begin{figure}[ht]
	\centering
	\includegraphics[width=.5\textwidth]{indiaMap-crop}
	\caption{This map illustrates the geographic distribution of all conflict centroids in India, according to the PRIO Conflict Site Dataset, and major cities from 1989 to 2007. }
	\label{fig:indiaMap}
\end{figure}

Nigeria and Cameroon have battled Boko Haram in a bloody conflict in their rural northern regions for half a decade. For the period between 2009 and 2013, the National Consortium for the Study of Terrorism and Responses to Terrorism (START) described Boko Haram as ``among the deadliest [terrorist groups] in the world'' \citep{pate:etal:2014}. The group's tactics range from car and suicide bombings to direct assaults to mass kidnappings. Nonetheless, Nigeria and Cameroon have both maintained strong economic growth during this period. The African Development Bank Group (AFDB) described Nigeria's economic growth from 2004 to 2014 as ``robust'' and projected ``moderate'' growth of 5\% for 2015 \citep{afdb:nigeria:2015}. On Cameroon, the AFDB said ``despite the security and humanitarian crisis in the region, Cameroonian growth remains strong at above 5\%'' \citep{afdb:cameroon:2015}.

It is important to note that Mexico does not appear in the civil war dataset chosen for this study. The UCDP definition of civil war requires that the conflict be fought over an ``incompatibility that concerns government and/or territory,'' a criteria not strictly met by the organized crime violence experienced by the country \citep[p.1]{themner:2014}. Nonetheless, we want to reiterate that this instance of armed conflict, of the magnitude typically associated with civil war, conforms to the expectations of our theory. Mexico, now nearly a decade into a violent and complicated conflict between several organized criminal enterprises and the federal government, has maintained healthy economic performance. For much of this time, the cartel violence generally occurred in rural areas along drug trafficking routes and not within major cities. \citet{beittel:2011} writes that ``drug trafficking-related killings remain concentrated in a relatively few cities.'' Meanwhile, several majors cities including the capital, Mexico City, have experienced relatively low levels of cartel violence.
